---
title: "Normalization_pc_18"
output: html_document
date: "2023-05-24"
---



```{r setup, include=FALSE, message=FALSE}
library(readxl)
library(tidyverse)
library(data.table)
#library(DT)
library(lmerTest)
library(stringr)
library(htmlwidgets)

```



#Constant
```{R}
file.name <- "pos_lipid"
outcomes <- c("WAZ" ,"WLZ" ,"LAZ" ,"HCAZ")
constant_IQR <- 1.5
cv_human_threshold <-0.1
```

#reading raw data
```{r}
raw_pc_18 <- read_csv(paste("C:/Users/priceade/Desktop/Fecal_Omics/Older Code and Workflows/processed_data_1/raw", file.name,".csv",sep = ""))
```

#Reading in PC 18 processed data
Assigning Compound Names Such as: P_LP.1, ...to Positive lipid metabolites
```{r}
pc_18_cmpd_info <- as.data.frame(read.table(file =paste("C:/Users/priceade/Desktop/Fecal_Omics/Older Code and Workflows/processed_data_1/cmpd_info_", file.name,".txt",sep = ""), header=T))
names(pc_18_cmpd_info)[names(pc_18_cmpd_info) == 'Compound'] <- 'Compound.Name'
fully_joined_pc18 <- as.data.frame(read.table(file =paste("C:/Users/priceade/Desktop/Fecal_Omics/Older Code and Workflows/processed_data_1/merged_file_", file.name,".txt",sep = ""), header=T))

Compound<- paste("P_LP.",1:nrow(pc_18_cmpd_info), sep = "")
pc_18_cmpd_info <- cbind(Compound,pc_18_cmpd_info)

#datatable(as.data.frame(pc_18_cmpd_info),rownames=F,extensions=c("Buttons","ColReorder"),
#          options=list(dom="Bfrtipl",scrollY=TRUE,scrollX=TRUE,colReorder=TRUE,buttons=c("excel","csv"),
#                       pageLength=10,lengthMenu=c(10,25,50)),
#          class="cell-border stripe", caption="Summary Information for Metabolites")

#write.csv(pc_18_cmpd_info, paste("compound_Information_", file.name,".csv",sep = ""), row.names=F)

## Naming Columns for "pc_18" 
colnames(fully_joined_pc18)[12:ncol(fully_joined_pc18)] <- Compound
fully_joined_pc18 <-  fully_joined_pc18[complete.cases(fully_joined_pc18[,Compound]),]
```

#check balancing

```{r}

check_balance <- cbind(fully_joined_pc18[,c(1:11)],fully_joined_pc18[,Compound])
check_balance$Type <- ifelse(str_detect(check_balance$StudyID, "pool"), "pool", "Human")
check_balance <- as.data.frame(check_balance[!(check_balance$Type=="pool"),])

batch_check <- check_balance%>%
  group_by(batch, Time, Group)%>%
summarise(n())

print(batch_check)

Time_check <- check_balance%>%
  group_by(Time)%>%
summarise(n())
print(Time_check)

Group_check <- check_balance%>%
  group_by(Group)%>%
summarise(n())
print(Group_check)

all_check <- check_balance%>%
  group_by(batch, Time, Group)%>%
summarise(n())
```

## Filter By Missingness

For this analysis, we will only consider metabolites seen in at least 25% of the plasma samples (regardless of time status). This will guarantee that the largest percent missing allowed is 75% across all compounds in our analysis.

```{r Missingness Exploration, echo = T, message = FALSE, results = FALSE, cache = F}
#how many compounds?

dim(raw_pc_18) #3298 compounds


dat_t <- raw_pc_18

#Subset to raw data only:
raw_dat <- dat_t

raw_human <- raw_dat[grep("pool", raw_dat$StudyID, invert = T),]
dim(raw_human) #167 samples

rownames(raw_human) <- raw_human$StudyID
raw_human <- raw_human[,-1]

#missingness by compounds
#turn into numeric
raw_human <- apply(raw_human, 2, as.numeric)

#colnames(raw_human) <- raw.mets$Compound
missing_values_counts <- apply(raw_human, 2, function(x) sum(x == 1))
missing_values_pcts <- apply(raw_human,2, function(x) sum(x==1)/167)

missing_df <- data.frame(Compound.Name = names(missing_values_counts),
                         Missing_Count = missing_values_counts,
                         Missing_Pct = missing_values_pcts)

```


#Filtering Missingness
```{R}

missing_df2 <- left_join(missing_df, pc_18_cmpd_info[,c(1,2)])
missing_df <- missing_df2[,c(4,1,2,3)]

missing_filtered <- merge(missing_df[missing_df$Missing_Pct<.9,], pc_18_cmpd_info[,c(1,2)], by = c("Compound","Compound.Name"), all.x=T)

Compound <- missing_filtered$Compound

length(Compound)
```


# Calculating Absence(missing) Values of Compounds at baseline in each diet:
```{r}
#Meat_base
pc_18_meat_base <- fully_joined_pc18[fully_joined_pc18$Group !="Dairy"&fully_joined_pc18$Time == "Base", ]
pc_18_dairy_base <- fully_joined_pc18[fully_joined_pc18$Group !="Meat"&fully_joined_pc18$Time == "Base", ]
## Calculating Missing in "Base", "Mid" and "End"

my_frame <- data.frame(matrix(ncol=2, nrow=2))
colnames(my_frame) <- c("Missing","Present")
rownames(my_frame) <- c("dairy","meat")

my_p_frame <- data.frame(matrix(ncol=1, nrow=length(Compound)))
colnames(my_p_frame) <- "P_value"
rownames(my_p_frame) <- Compound


for (i in 1: length(Compound)){
  my_frame[1,1] <- (sum(pc_18_dairy_base[,Compound[i]]==1, na.rm = T))
  my_frame[1,2] <- (sum(pc_18_dairy_base[,Compound[i]]!=1, na.rm = T))
  my_frame[2,1] <- (sum(pc_18_meat_base[,Compound[i]]==1, na.rm = T))
  my_frame[2,2] <- (sum(pc_18_meat_base[,Compound[i]]!=1, na.rm = T))
  fisher_i <- fisher.test(my_frame)
  my_p_frame [i,1]<- fisher_i$p.value
}


paste("We have significant evidence that there is a difference between the compounds present at baseline in Dairy and Meat diet for",length(my_p_frame[my_p_frame$P_value< 0.05,]),"out of",nrow(my_p_frame),"compounds", sep = " ")
```

#Batch Correction preparation
```{r}

pc_18_dat <- cbind(fully_joined_pc18[,c(1:11)],fully_joined_pc18[,Compound])

pc_18_dat[,c("ms.vial", "Group", "Time")] <- lapply(pc_18_dat[,c("ms.vial", "Group", "Time")],factor)
pc_18_dat[,outcomes] <- lapply(pc_18_dat[,outcomes],as.numeric)
pc_18_dat[,Compound] <- lapply(pc_18_dat[,Compound],as.numeric)
pc_18_dat[,c("batch", "count")] <- lapply(pc_18_dat[,c("batch", "count")],as.numeric)

## Adding a New Column to Identify QC Values from Human
pc_18_dat$Type <- ifelse(str_detect(pc_18_dat$StudyID, "pool"), "pool", "Human")
```

```{R}
## Removing QC Rows
pc_18_dat <- as.data.frame(pc_18_dat[!(pc_18_dat$Type=="pool"),])

pc_18_dat <- droplevels(pc_18_dat)

#str(log2_phillic)
```


# Outliers and Correction
Identifying outlier points
```{r}
## Identifying and Removing Outliers
## replacing all 0 values in Metabolite with NAs 
pc_18_dat[,Compound][pc_18_dat[,Compound]==1] <- NA

## Identifying outliers
cmpd_outlier <- as.data.frame(matrix(nrow = nrow(pc_18_dat), ncol = ncol(pc_18_dat[,Compound])))
for(i in 1:length(Compound)){
  cmpd_loop <- pc_18_dat[,c(Compound[i], "count","batch")]
  names(cmpd_loop) <- c("Compound", "count", "batch")
  linear_reg <- lm(Compound~count*batch, data=cmpd_loop) 
  cmpd_residuals <- as.vector(linear_reg$residuals)
  cmpd_q3 <- quantile(cmpd_residuals, 0.75)
  cmpd_q1 <- quantile(cmpd_residuals, 0.25)
  iqr <- cmpd_q3 - cmpd_q1
  cmpd_outlier[names(residuals(linear_reg)),i] <- cmpd_residuals 
  cmpd_outlier[,i] <- ifelse(cmpd_outlier[,i] > (cmpd_q3 + constant_IQR*iqr) | cmpd_outlier[,i] < (cmpd_q1 - constant_IQR*iqr),F,T)
}

## Replacing T and F with 0 and 1, and Removing Empty Rows
cmpd_outlier <- as.data.frame(sapply(cmpd_outlier, as.numeric))
cmpd_outlier <- cmpd_outlier[rowSums(is.na(cmpd_outlier))!=ncol(cmpd_outlier),]
      
## Removing Outliers 
updated_cmpd <- pc_18_dat
updated_cmpd[,Compound] <- mapply('*', updated_cmpd[,Compound], cmpd_outlier)
updated_cmpd[,Compound][updated_cmpd[,Compound] == 0] <- NA #setting outliers to NA
```


## Normalization
Normalizing the Dataset After Removing Outliers
```{r}

#str(updated_cmpd)


## Calculating the normalized values
cmpd_normalized <- as.data.frame(matrix(nrow = nrow(updated_cmpd), ncol = ncol(updated_cmpd[,Compound])))

predicted_value <- as.data.frame(matrix(nrow = nrow(updated_cmpd), ncol = ncol(updated_cmpd[,Compound])))

#unique(updated_cmpd$count)
#str(updated_cmpd[,c(1:12)])

for(i in 1:length(Compound)){
  linear_reg_normalized <- lm(updated_cmpd[,Compound[i]]~count*batch, data=updated_cmpd)
  predicted_reg <- predict.lm(linear_reg_normalized, newdata = pc_18_dat)
  predicted_value[,i]<-as.data.frame(predicted_reg)
  cmpd_normalized[,i] <- as.data.frame(pc_18_dat[,Compound[i]]-predicted_reg)
}

colnames(cmpd_normalized) <- Compound
colnames(predicted_value) <- Compound

cmpd_normalized <- cbind(updated_cmpd[,c(1:11)],cmpd_normalized)
cmpd_predicted <- cbind(updated_cmpd[,c(1:11)],predicted_value)
```

```{r}
#Meat_base
pc_18_meat_base2 <- cmpd_normalized[cmpd_normalized$Group !="Dairy"&cmpd_normalized$Time == "Base", ]
pc_18_dairy_base2 <- cmpd_normalized[cmpd_normalized$Group !="Meat"&cmpd_normalized$Time == "Base", ]
## Calculating Missing in "Base", "Mid" and "End"

my_frame <- data.frame(matrix(ncol=2, nrow=2))
colnames(my_frame) <- c("Missing","Present")
rownames(my_frame) <- c("dairy","meat")

my_p_frame <- data.frame(matrix(ncol=1, nrow=length(Compound)))
colnames(my_p_frame) <- "P_value"
rownames(my_p_frame) <- Compound


for (i in 1: length(Compound)){
  my_frame[1,1] <- (sum(is.na(pc_18_dairy_base2[,Compound[i]])))
  my_frame[1,2] <- (sum(!is.na(pc_18_dairy_base2[,Compound[i]])))
  my_frame[2,1] <- (sum(is.na(pc_18_meat_base2[,Compound[i]])))
  my_frame[2,2] <- (sum(!is.na(pc_18_meat_base2[,Compound[i]])))
  fisher_i <- fisher.test(my_frame)
  my_p_frame [i,1]<- fisher_i$p.value
}


pc_18_dairy_base2[,is.na(Compound[i])]
paste("We have significant evidence that there is a difference between the compounds present at baseline in Dairy and Meat diet for",length(my_p_frame[my_p_frame$P_value< 0.05,]),"out of",nrow(my_p_frame),"compounds", sep = " ")
```

#minimum computing
```{r}
cmpd_minimum <- as.data.frame(matrix(nrow = nrow(updated_cmpd), ncol = ncol(updated_cmpd[,Compound])))
# i<-1
for(i in 1:length(Compound)){
  raw_min <- min(pc_18_dat[,Compound[i]], na.rm = T)
  mod_min <- min(cmpd_normalized[,Compound[i]]-cmpd_predicted[,Compound[i]],na.rm = T)
  cmpd_min <- raw_min - mod_min
  cmpd_minimum[,i]<- cmpd_min
  }
colnames(cmpd_minimum) <- colnames(updated_cmpd[,Compound])

cmpd_normalized_shift2<- cmpd_normalized[,Compound]+cmpd_minimum

cmpd_normalized_shift <-cbind(updated_cmpd[,c(1:11)],cmpd_normalized_shift2)

#write.csv(cmpd_normalized_shift, paste("shiffted_", file.name,".csv",sep = ""), row.names=F)

```

<!-- #shifting -->
<!-- ```{r} -->
<!-- min_shift <- apply(pc_18_dat[Compound],2,min,na.rm=T) -->
<!-- cmpd_normalized_shift <- cmpd_normalized -->
<!-- for(i in 1:length(Compound)){ -->
<!--   cmpd_normalized_shift[Compound[i]] <- cmpd_normalized_shift[Compound[i]]+min_shift[i] -->
<!-- } -->
<!-- #write.csv(cmpd_normalized_shift, paste("./data_table/normalized_", file.name,".csv",sep = ""), row.names=F) -->
<!-- ``` -->




#remove low variability. Only keep cpds >10% missing or non-0 cv > 0.1
```{r CV Filter, message=F, warning=F, results=T, echo=F}
#Function to calculate CV given a vector of values
cv_calc <- function(x){
  x <- x[complete.cases(x)]
  mu <- mean(x)
  sig <- sd(x)
  
  cv <- abs(sig/mu)
  return(cv)
}
```



## Filter by Coefficient of Variation in human samples and missingness threshold

```{r}
#Create a data set consisting of just samples
cmpd_normalized_shift$Type <- ifelse(str_detect(cmpd_normalized_shift$StudyID, "pool"), "pool", "Human")

Human_dat <- cmpd_normalized_shift[cmpd_normalized_shift$Type == "Human",Compound]

#Human_dat[is.na(Human_dat)] <- 0

#apply the CV function by metabolite
cv_Human_dat <- data.frame(Compound = Compound,
                     CV_value = apply(Human_dat[,Compound], 2, function(x) cv_calc(x)))

summary(cv_Human_dat$CV_value)
hist(cv_Human_dat$CV_value)

missing.tmp <- missing_df2

human_filtering_data <- left_join(cv_Human_dat,missing.tmp)

hist(human_filtering_data$Missing_Pct)
#Filter by CV threshold


#Initial data
#3296

more_10_missing <- human_filtering_data%>%
  filter(Missing_Pct>0.10)

cv_human_threshold <- quantile(cv_Human_dat$CV_value, .02)
#1396

coef_var_high <- human_filtering_data%>%
  filter(CV_value>cv_human_threshold)
# 314
2722
#30
#247
#1769

 #threshold: 1 ->#65
#threshold: 0.5 ->#386
 #threshold: 0.1 ->#1572
 #threshold: 0.1 ->#1769

Compound_2 <- human_filtering_data[human_filtering_data$CV_value>cv_human_threshold | human_filtering_data$Missing_Pct > 0.10, "Compound"]

print(paste("After using a threshold of ", cv_human_threshold, " to remove compounds with low variability  for human samples there are ", length(Compound_2)," compounds remaining.", sep = ""))
```

```{r}
cmpd_normalized_shift <-cbind(updated_cmpd[,c(1:11)],cmpd_normalized_shift[,Compound_2])
#pos_lip <- cmpd_normalized_shift

cmpd_normalized_shift <- cmpd_normalized_shift %>%
  mutate(across(where(is.numeric),function(x) {if_else(is.na(x),min(x,na.rm=T),x)}))



#we are just assessing base/end for now
cmpd_normalized_shift_end <- cmpd_normalized_shift[cmpd_normalized_shift$Time == 'End',]


########################################Explore significance############################
micro_dairy <- cmpd_normalized_shift_end[cmpd_normalized_shift_end$Group == 'Dairy',]
micro_meat <- cmpd_normalized_shift_end[cmpd_normalized_shift_end$Group == 'Meat',]

micro_d_data <- as.matrix(micro_dairy[,12:3244])
class(micro_d_data) <- "numeric"
micro_m_data <- as.matrix(micro_meat[,12:3244])
class(micro_m_data) <- "numeric"

print(paste0('metabolites: dairy, pos lip ', dim(micro_d_data)[1]))
print(paste0('metabolites: meat, pos lip', dim(micro_m_data)[1]))

micro_d_data <- log2(micro_d_data)
micro_m_data <- log2(micro_m_data)

p_val <- c()
stat <- c()

for (i in 1:dim(micro_d_data)[2]){
    test <-wilcox.test(micro_d_data[,(i)],micro_m_data[,(i)], paired = FALSE, exact = FALSE)
    p_val <- append(p_val,as.numeric(test$p.value))
    stat <- append(stat,as.numeric(test$statistic))
  }
bon_cor <- .05/dim(micro_d_data)[2]
fdr_cor <- p.adjust(p_val, method = 'fdr', n= dim(micro_d_data)[2])

###############################################################################################################################################

#calculate the mean of each feature val per samp
dairy = as.vector(apply(micro_d_data, 2, mean))

#calcuate the mean of each gene per test group
meat = as.vector(apply(micro_m_data, 2, mean))

fold_change <- -1*(dairy-meat)
full_results_pos_lip <- cbind.data.frame('feature'=colnames(micro_d_data), 'p_val' = p_val, 'fdr' = fdr_cor, 'fold_change' = fold_change)

#merge fold change info with key for features
#key <- read.csv("C://Users/priceade/Desktop/Fecal_Omics/Older Code and Workflows/All_micro_metab_KEY.csv")
#full_results_neg_aq_key <- merge(full_results_neg_aq, key, by.x = "feature", by.y = "code", all.x =TRUE)
#write.csv(full_results_neg_aq_key, "C:/Users/priceade/Desktop/Fecal_Omics/Current_Code&Workflow/Final_Manuscript_Figures/diff_abun_neg_aq.csv", row.names=FALSE)



###############################################################################################################################################



########################################Explore significance base/end############################
#we are just assessing base/end for now
cmpd_normalized_shift <- cmpd_normalized_shift[cmpd_normalized_shift$Time == 'End' | cmpd_normalized_shift$Time == 'Base',]

micro_labeled <- cmpd_normalized_shift[cmpd_normalized_shift$Vial.ID %in% names(which(table(cmpd_normalized_shift$Vial.ID) > 1)), ]

micro_base <- micro_labeled[micro_labeled$Time == 'Base',]
micro_base <- micro_base[order(micro_base$Vial.ID),]
micro_end <- micro_labeled[micro_labeled$Time == 'End',]
micro_end <- micro_end[order(micro_end$Vial.ID),]


print(paste0('metabolites: base, pos lip ', dim(micro_base)[1]))
print(paste0('metabolites: end, pos lip ', dim(micro_end)[1]))


micro_b_data <- as.matrix(micro_base[,12:3244])
class(micro_b_data) <- "numeric"
micro_e_data <- as.matrix(micro_end[,12:3244])
class(micro_e_data) <- "numeric"

micro_b_data <- log2(micro_b_data)
micro_e_data <- log2(micro_e_data)

p_val <- c()
stat <- c()

for (i in 1:dim(micro_b_data)[2]){
    test <-wilcox.test(micro_b_data[,(i)],micro_e_data[,(i)], paired = TRUE, exact = FALSE)
    p_val <- append(p_val,as.numeric(test$p.value))
    stat <- append(stat,as.numeric(test$statistic))
  }



############################################################################################################################################

#calculate the mean of each feature val per samp
base = as.vector(apply(micro_b_data, 2, mean))

#calcuate the mean of each gene per test group
end = as.vector(apply(micro_e_data, 2, mean))

fold_change <- -1*(base-end)




full_results_pos_lip_time <- cbind.data.frame('feature'=colnames(micro_b_data), 'p_val' = p_val, 'fdr' = fdr_cor, 'fold_change' = fold_change)
############################################################################################################################################


```








```{r}
meat_dairy_res <- rbind.data.frame(full_results_neg_aq, full_results_neg_lip, full_results_pos_lip, full_results_pos_aq)
meat_dairy_res$fdr_cor <- p.adjust(meat_dairy_res$p_val, method = 'fdr', n= dim(meat_dairy_res)[1])




key <- read.csv("C://Users/priceade/Desktop/Fecal_Omics/Older Code and Workflows/All_micro_metab_KEY.csv")
meat_dairy_key <- merge(meat_dairy_res, key, by.x = "feature", by.y = "code", all.x =TRUE)
#write.csv(meat_dairy_key, "C://Users/priceade/Desktop/Fecal_Omics/Current_Code&Workflow/Final_Manuscript_Figures/diff_abun_metabolites_MEAT_DAIRY.csv", row.names=FALSE)



base_end_res <- rbind.data.frame(full_results_neg_aq_time, full_results_neg_lip_time, full_results_pos_lip_time, full_results_pos_aq_time)
base_end_res$fdr_cor <- p.adjust(base_end_res$p_val, method = 'fdr', n= dim(base_end_res)[1])

#merge fold change info with key for features
key <- read.csv("C://Users/priceade/Desktop/Fecal_Omics/Older Code and Workflows/All_micro_metab_KEY.csv")
base_end_key <- merge(base_end_res, key, by.x = "feature", by.y = "code", all.x =TRUE)
#write.csv(base_end_key, "C://Users/priceade/Desktop/Fecal_Omics/Current_Code&Workflow/Final_Manuscript_Figures/diff_abun_metabolites_BASE_END.csv", row.names=FALSE)





```





```{r}


#base_end plot
#bonf
col_meta <- ifelse(((-1*log10(base_end_res$fdr_cor) > (-1*log10(.05))) & base_end_res$fold_change <0), "burlywood2", 
                   ifelse(((-1*log10(base_end_res$fdr_cor) > (-1*log10(.05))) & base_end_res$fold_change >0), "lightblue", "grey"))

col_meta <- ifelse((base_end_res$p_val <= .05/length(base_end_res$p_val) & base_end_res$fold_change <0), "darkorange4", 
                   ifelse((base_end_res$p_val <= .05/length(base_end_res$p_val)  & base_end_res$fold_change >0), "darkblue", col_meta))
###################visualize dis of sig w/ large fold change#####################
p_vals_filtered <- base_end_res[base_end_res$fdr_cor > .05,]

library(ggplot2)
ggplot(data = base_end_res, aes(x = fold_change, y = -1*log10(p_val))) + 
  geom_point(colour=col_meta) + ggtitle("Metabolome: Baseline v. Endpoint") +
  geom_hline(yintercept=-1*log10(min(p_vals_filtered$p_val)), linetype='solid', col = 'darkgrey') +
  geom_hline(yintercept=-1*log10(.05/length(base_end_res$p_val)), linetype='solid', col = 'black') +
  xlab('log2 fold change') +ylab('-log10(p-values)') +
  theme(axis.line = element_line(colour = "black"), 
        panel.border = element_rect(color = "black", fill = NA, size = 1), 
        axis.text=element_text(size=20),
        axis.title=element_text(size=20,face="bold"),
        plot.title = element_text(size = 20, face = "bold", hjust = 0.5)) 





#bonf
col_meta <- ifelse(((-1*log10(meat_dairy_res$fdr_cor) > (-1*log10(.05))) & meat_dairy_res$fold_change <0), "burlywood2", 
                   ifelse(((-1*log10(meat_dairy_res$fdr_cor) > (-1*log10(.05))) & meat_dairy_res$fold_change >0), "lightblue", "grey"))

col_meta <- ifelse((meat_dairy_res$p_val <= .05/length(meat_dairy_res$p_val) & meat_dairy_res$fold_change <0), "darkorange4", 
                   ifelse((meat_dairy_res$p_val <= .05/length(meat_dairy_res$p_val)  & meat_dairy_res$fold_change >0), "darkblue", col_meta))

###################visualize dis of sig w/ large fold change#####################
p_vals_filtered <- meat_dairy_res[meat_dairy_res$fdr > .05,]

library(ggplot2)
ggplot(data = meat_dairy_res, aes(x = fold_change, y = -1*log10(p_val))) + 
  geom_point(colour=col_meta) + ggtitle("Metabolome: Dairy v. Meat") +
  geom_hline(yintercept=-1*log10(min(p_vals_filtered$p_val)), linetype='solid', col = 'darkgrey') +
  xlab('log2 fold change') +ylab('-log10(p-values)') +
  theme(axis.line = element_line(colour = "black"), 
        panel.border = element_rect(color = "black", fill = NA, size = 1), 
        axis.text=element_text(size=20),
        axis.title=element_text(size=20,face="bold"),
        plot.title = element_text(size = 20, face = "bold", hjust = 0.5)) 



```













#Ranking 
```{R}
cmpd_normalized_shift <-cbind(updated_cmpd[,c(1:11)],cmpd_normalized_shift[,Compound_2])


## Replacing "NA" with -999 for Ranking

# str(cmpd_updated)
# unique(cmpd_updated$Group)
# unique(cmpd_updated$Time)

#Ranking dataset

cmpd_ranked <- cmpd_normalized_shift
cmpd_ranked[,Compound_2] <- as.data.frame(apply(cmpd_ranked[,Compound_2],2,function(x)
 rank(replace(x,x==0, NA), na='keep')))
cmpd_ranked[,Compound_2] <- as.data.frame(apply(cmpd_ranked[,Compound_2],2,function(x)
 replace(x,is.na(x), 0)))
#length(unique(cmpd_ranked$Vial.ID))

cmpd_ranked$Vial.ID <- as.factor(cmpd_ranked$Vial.ID)

#Cmpd Ranked
#write.csv(cmpd_ranked, paste("ranked_", file.name,".csv",sep = ""), row.names=F)

```

#Rescale data
```{r}
cmpd_rescaled <- as.data.frame(matrix(nrow = nrow(cmpd_ranked), ncol = ncol(cmpd_ranked[,Compound_2])))
for(i in 1:length(Compound_2)){
  cmpd_rescaled[,i]<- cmpd_ranked[,Compound_2[i]]/max(cmpd_ranked[,Compound_2[i]])
  }
colnames(cmpd_rescaled) <- Compound_2
cmpd_rescaled <-cbind(cmpd_ranked[,c(1:11)],cmpd_rescaled)

dim(cmpd_rescaled)
#Cmpd Rescaled
#write.csv(cmpd_rescaled, paste("rescaled_", file.name,".csv",sep = ""), row.names=F)

```

#Put into format for Pecanpy
#Format data into 3 columns (Baby, Compound, Rank) (Both normalized shifted, ranked) 
##Positive Lipid
```{r}
ranked_pos_lipid <- cmpd_rescaled
ranked_pos_lipid$sample.name <- paste(ranked_pos_lipid$Vial.ID,ranked_pos_lipid$Time,sep="_")
Information_code_pos_lipid <- ranked_pos_lipid[,c("sample.name","Group","Time","ms.vial","Vial.ID","count","batch","WAZ","WLZ","LAZ","HCAZ")]


transformed_data_pos_lipid<- ranked_pos_lipid[,c("sample.name", Compound_2)]
#write.csv(transformed_data_pos_lipid, paste("C:/Users/priceade/Desktop/Fecal_Omics/processed_data_1/Regression Model Data/", file.name,".csv",sep = ""), row.names=F)

long_data_pos_lipid <- transformed_data_pos_lipid%>%
  pivot_longer(starts_with("P_LP."), names_to = "Compound", values_to = "Norm_value")

#Long_compd
#number of unique baby 
length(unique(long_data_pos_lipid$sample.name))
#167

#number of unique compounds 
length(unique(long_data_pos_lipid$Compound))
#2789

long_data_pos_lipid <- long_data_pos_lipid[long_data_pos_lipid$Norm_value >0,]

write_tsv(long_data_pos_lipid, paste("C:/Users/priceade/Desktop/Fecal_Omics/Current_Code&Workflow/Data/long_format_", file.name,".tsv",sep = ""))

```

Check

```{R}

compound.1.init <- pc_18_dat[,"P_LP.10"]

compound.1.no.outlier <- updated_cmpd[,"P_LP.10"]

compound.1.predicted <- cmpd_predicted[,"P_LP.10"]

compound.1.norm <-cmpd_normalized[,"P_LP.10"]

compound.1.norm_shiffted <- cmpd_normalized_shift[,"P_LP.10"]

compound.1.ranked <- cmpd_ranked[,"P_LP.10"]

compound.1.rescaled <- cmpd_rescaled[,"P_LP.10"]


data_check <- data.frame(cbind(pc_18_dat[,"ms.vial"],compound.1.init,compound.1.no.outlier,compound.1.predicted,compound.1.norm,compound.1.norm_shiffted,compound.1.ranked, compound.1.rescaled))

long_data <-  data_check%>%
  pivot_longer(!V1, names_to = "data_type", values_to = "value")


long_data$data_type <- factor(long_data$data_type, levels=c("compound.1.init","compound.1.no.outlier","compound.1.predicted","compound.1.norm","compound.1.norm_shiffted","compound.1.ranked", "compound.1.rescaled"))

levels(long_data$data_type)


ggplot(long_data, aes(x = value, y = data_type, fill = data_type)) +
  geom_density_ridges(
    aes(point_color = data_type, point_fill = data_type),
    alpha = .2, point_alpha = 1, jittered_points = TRUE
  ) +
  scale_point_color_hue(l = 40)+
  theme_classic()

```